---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# Genetic Algorithm for Generalized Additive Models

<!-- badges: start -->
<!-- badges: end -->

This package implements a genetic algorithm which performs simulatenous variable selection and structure discovery in generalized additive models. For a given dependent variable and a set of explanatory variables, the genetic algorithm determines which regressors should be included linearly, which nonparametrically, and which should be excluded from the regression equation. The aim is to minimize the Bayesian Information Criterion value of the model.

## Installation

You can install the released version of gagam from GitHub with:

``` r
library(devtools)
install_github("markcus1/gagam")
```

## Example

This is a basic example which shows you how to use the package for variable selection and structure discovery:

```{r example}
library(gagam)
## basic example code
N <- 500
set.seed(123)
xdat <- matrix(rnorm(N*10,0,1),nrow=N,ncol=10)
ydat <- 4*xdat[,1]+5*xdat[,2]+6*xdat[,3]+(xdat[,4])^2 + 4 + rnorm(N,0,0.25)
example_gagam <- gagam(ydat,xdat,Kvar = 6,no_gen = 50)
example_gagam
```

## Further Information

The help pages and the package manual contain more information on the parameters one can specify.

For details on the algorithm see PAPER.
